Markdown to estimate the abundance of brain cell types with the reprocessed counts from Mayo cohorts (with TMM-VOOM normalisation)

### 2017-02-03 code/Rmd/Mayo_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd
Added markdown to deconvolute celltype fraction from Mayo data

---
title: "Markdown to estimate the abundance of brain cell types with the reprocessed counts from ROSMAP cohort (with TMM-VOOM normalisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse)

@@

knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd",
                                 parentId = "syn5570291",
                                 entityName = 'ROSMAP Cell Type Abundance with Reprocessed RNASeq')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5570291';
activityName = 'Cell type abundance in ROSMAP';
activityDescription = 'Cell type abundance in ROSMAP with TMM-VOOM normalised reprocessed RNASeq counts';

thisFileName <- 'ROSMAP_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='celltype')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download expression data
COUNT_ID <- 'syn7750270';
ALL_USED_IDs <- COUNT_ID
COUNT_OBJ <- synGet(COUNT_ID)
COUNT <- read.table(COUNT_OBJ@filePath,header=T,sep='\t',check.names = F)

# Get clinical metadata
METADATA.CLINICAL_ID <- 'syn3191087'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.CLINICAL_ID
METADATA.CLINICAL_OBJ <- synGet(METADATA.CLINICAL_ID)
METADATA.CLINICAL <- read.table(METADATA.CLINICAL_OBJ@filePath,sep=',',header=T)

# Get technical covariates
METADATA.TECH_ID <- 'syn4300313'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.TECH_ID
METADATA.TECH_OBJ <- synGet(METADATA.TECH_ID)
METADATA.TECH <- read.table(METADATA.TECH_OBJ@filePath,sep='\t',header=T)

# Fix error in technical covariates data
KEY_ID <- 'syn3382527'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = KEY_ID  
KEY <- read.csv(synGet(KEY_ID)@filePath) %>% 
  filter(mrna_data == 1) %>%
  dplyr::select(projid, mrna_id) %>%
  separate(mrna_id, c('a','b','batch'), sep = '_') %>% 
  unite(Sampleid, a, b) %>%
  dplyr::select(-batch) %>%
  unique

# Match technical and clinical covariates
METADATA <- METADATA.TECH %>% 
  dplyr::select(-projid) %>%
  dplyr::left_join(KEY) %>%
  dplyr::left_join(METADATA.CLINICAL)

# Pick higher quality RIN batch for sample 492_120515
METADATA <- METADATA %>%
  group_by(Sampleid) %>%
  top_n(1, RINcontinuous)
```

### Data preprocessing
```{r preprocess.data,cache=TRUE, echo=TRUE}
# Remove samples with no cogdx, RIN and PMI scores
METADATA <- METADATA %>%
  ungroup %>%
  filter(Sampleid %in% colnames(COUNT)) %>%
  filter(!is.na(cogdx)) %>%
  filter(!is.na(RINcontinuous)) %>%
  filter(!is.na(pmi)) %>%
  filter(!is.na(Sampleid)) %>%
  filter(!(Sampleid %in% c('380_120503'))) %>%
  as.data.frame()
```

```{r preprocess.data1,cache=TRUE}
# Match covariates to expression data
indToRetain = intersect(METADATA$Sampleid, colnames(COUNT))
removedIDs = setdiff(colnames(COUNT), METADATA$Sampleid)

COUNT = COUNT[,indToRetain]

rownames(METADATA) = METADATA$Sampleid
METADATA = METADATA[indToRetain,]
```
Dorsolateral prefrontal cortex of `r dim(COUNT)[2]` subjects from the ROS and MAP cohorts are used for the analysis. Following sample are removed `r paste(removedIDs, collapse = ',')`

### Get Covariates
```{r covariates.clustering, cache=TRUE}
#"braaksc","ceradsc","cts_mmse30_first_ad_dx","cts_mmse30_lv"
FactorCovariates <- c("Batch", "msex", "race", "spanish", "cogdx")
ContCovariates <- c("RINcontinuous","age_first_ad_dx","age_death","age_at_visit_max","pmi","educ")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
COVARIATES <- data.frame(lapply(COVARIATES,function(x){x <- sapply(x,function(y){str_replace_all(as.character(y),'\\+','')})}))
rownames(COVARIATES) <- METADATA$Sampleid

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)

# Add in RIN^2 values
COVARIATES$RINcontinuous2 = COVARIATES$RINcontinuous^2
ContCovariates = c(ContCovariates, 'RINcontinuous2')
```

### CPM Normalisation
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

Remove genes that have less than 1 cpm counts in at least 50% of samples per cogdx
```{r cpmnormalisation}
genesToAnalyze = COVARIATES %>%
  rownameToFirstColumn('Sampleid') %>%
  dlply(.(cogdx), .fun = function(mtd, count){
  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$Sampleid],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)
genesToAnalyze = unlist(genesToAnalyze) %>% unique()

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples

### Normalisation (with NULL)
Initial normalisation usign voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# TMM normalisation
TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix, method = 'TMM')

@@

# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=NULL, plot=T)
```

### Get cell type specific signatures and expression data from synapse
```{r get.signatures}
ALL_USED_IDs = c(ALL_USED_IDs, 'syn8068511', 'syn8068475')
# Signatures from Zhang,Y et al., 2014
sig.zhang = downloadFile('syn8068511') %>%
  dlply(.(Set), .fun = function(x){return(unique(x$Gene))})

@@

```{r get.gene.symbols, cache=TRUE, eval=T}
backgroundGenes = data.frame(gene.id = rownames(VOOM.GENE_EXPRESSION$E)) %>%
  dplyr::mutate(ensembl_gene_id = gene.id) %>%
  tidyr::separate(ensembl_gene_id, c('ensembl_gene_id', 'position'), sep = '\\.')

@@

write.table(ref.expr1, file = 'ref.expr.txt', sep = '\t', quote=F, row.names = F)
write.table(EXPR1[,1:20], file = 'mixture.expr.txt', sep = '\t', quote=F, row.names = F)

@@

p = ggplot(cellprop1, aes(x = SampleID, y = Fraction, fill = Celltype, color = cogdx)) + geom_bar(stat = 'identity')
p = p + theme(axis.text.x = element_text(angle = 90, hjust = 1))
# p = p + facet_grid(.~cogdx)
p

p = ggplot(cellprop1, aes(x = Celltype, y = Fraction, color = cogdx)) + geom_boxplot()
p = p + theme(axis.text.x = element_text(angle = 90, hjust = 1))
p
```
### Synapse store
```{r syn.store, cache=TRUE}
CODE = Folder(name = 'ROSMAP Cell Type Abundance with Reprocessed RNASeq', parentId = parentId)
CODE = synStore(CODE)

cellprop %>%
  rownameToFirstColumn('SampleID') %>%
  write.table(file = 'ROSMAP_CellFraction_Reproc.tsv', row.names = F, sep = '\t', quote = F)
obj = File('ROSMAP_CellFraction_Reproc.tsv', name = 'ROSMAP Cell Fraction', parentId = CODE$properties$id)
obj = synStore(obj, activityName = activityName, activityDescription = activityDescription,
               used = ALL_USED_IDs, executed = thisFile)
```

>>>>

---
title: "Markdown to estimate the abundance of brain cell types with the reprocessed counts from Mayo cohorts (with TMM-VOOM normalisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse)

@@

knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd",
                                 parentId = "syn5570325",
                                 entityName = 'MAYO Cell Type Abundance with Reprocessed RNASeq')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5570325';
activityName = 'Cell type abundance in Mayo';
activityDescription = 'Cell type abundance in Mayo with TMM-VOOM normalised reprocessed RNASeq counts';

thisFileName <- 'Mayo_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='celltype')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Get reprocessed sample counts for temporal cortex
COUNT_TC_ID <- 'syn7987765';
ALL_USED_IDs <- COUNT_TC_ID
COUNT_TC <- read.table(synGet(COUNT_TC_ID)@filePath, header=T, sep='\t', check.names = F)
colnames(COUNT_TC) = paste(colnames(COUNT_TC), 'TCX', sep = '_')

# Get reprocessed sample counts for cerebellum
COUNT_CE_ID <- 'syn7513807';
ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_CE_ID
COUNT_CE <- read.table(synGet(COUNT_CE_ID)@filePath, header=T, sep='\t', check.names = F)

COUNT = cbind(COUNT_TC, COUNT_CE[rownames(COUNT_TC),])

# Get clinical metadata
METADATA_TC_ID <- 'syn3817650'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_TC_ID
METADATA_TC <- read.table(synGet(METADATA_TC_ID)@filePath,sep='\t',header=T, stringsAsFactors=F) %>%
  tidyr::separate(ID, c('Donor_ID', 'BrainRegion'), sep = '_') %>%
  dplyr::mutate(ID = paste(Donor_ID, BrainRegion, sep = '_'))

# Get clinical metadata
METADATA_CE_ID <- 'syn5223705'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_CE_ID
METADATA_CE <- read.table(synGet(METADATA_CE_ID)@filePath,sep='\t',header=T, stringsAsFactors=F) %>%
  tidyr::separate(SampleID, c('Donor_ID', 'BrainRegion')) %>%
  dplyr::mutate(ID = paste(Donor_ID, BrainRegion, sep = '_')) %>%
  dplyr::rename(Gender = Sex, FLOWCELL = Flowcell)

METADATA = rbindlist(list(METADATA_TC, METADATA_CE), use.names = T, fill = T) %>%
  as.data.frame
```

### Data preprocessing
```{r preprocess.data,cache=FALSE, echo=TRUE}
# Fix metadata
METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
METADATA$Diagnosis = gsub("Pathologic Aging","Pathologic_Aging", METADATA$Diagnosis)
METADATA$BrainRegion.Diagnosis = paste(METADATA$BrainRegion, METADATA$Diagnosis, sep = '.')
METADATA$RIN2 = METADATA$RIN^2
METADATA <- METADATA %>%
  filter(!is.na(Source)) %>%
  dplyr::filter(!(ID %in% c("11311_CER", "1923_CER", "1923_TCX", 
                            "11396_TCX", "11294_TCX", "11408_TCX")))
```
Following samples were considered outliers and removed based on PCA and/or eucledian distance: 11311\_CER, 1923\_CER, 1923\_TCX, 11396\_TCX, 11294\_TCX, 11408\_TCX
```{r preprocess.data1}
# Match covariates to expression data
indToRetain = intersect(METADATA$ID, colnames(COUNT))
indRemoved = setdiff(colnames(COUNT), METADATA$ID)

COUNT <- COUNT[, indToRetain]

rownames(METADATA) = METADATA$ID
METADATA = METADATA[indToRetain,]
```
`r dim(COUNT)[2]` samples from `r length(unique(METADATA$Donor_ID))` subjects were obtained from the Mayo cohorts in AMP-AD reprocessed rnaseq project.

Following samples were removed `r indRemoved`

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
rownames(COVARIATES) <- METADATA$ID

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
```
### Gene Filtering
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

Remove genes that have less than 1 cpm counts in at least 50% of samples per BrainRegion.Diagnosis
```{r cpmnormalisation}
genesToAnalyze = COVARIATES %>%
  rownameToFirstColumn('Sampleid') %>%
  dlply(.(BrainRegion.Diagnosis), .fun = function(mtd, count){
    processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$Sampleid],
                                                     MIN_GENE_CPM=1, 
                                                     MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
    processed.counts$filteredExprMatrix$genes
    }, COUNT)
genesToAnalyze = unlist(genesToAnalyze) %>% unique()

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples
### Normalisation (with NULL)
Initial normalisation usign voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# TMM normalisation
TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix, method = 'TMM')

@@

# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=NULL, plot=T)
```
### Get cell type specific signatures and expression data from synapse
```{r get.signatures}
ALL_USED_IDs = c(ALL_USED_IDs, 'syn8068511', 'syn8068475')
# Signatures from Zhang,Y et al., 2014
sig.zhang = downloadFile('syn8068511') %>%
  dlply(.(Set), .fun = function(x){return(unique(x$Gene))})

@@

```{r get.gene.symbols, cache=TRUE, include=FALSE}
backgroundGenes = data.frame(gene.id = rownames(VOOM.GENE_EXPRESSION$E)) %>%
  dplyr::mutate(ensembl_gene_id = gene.id) %>%
  tidyr::separate(ensembl_gene_id, c('ensembl_gene_id', 'position'), sep = '\\.')

@@

write.table(ref.expr1, file = 'ref.expr.txt', sep = '\t', quote=F, row.names = F)
write.table(EXPR1, file = 'mixture.expr.txt', sep = '\t', quote=F, row.names = F)

@@

p = ggplot(cellprop1, aes(x = SampleID, y = Fraction, fill = Celltype, color = cogdx)) + geom_bar(stat = 'identity')
p = p + theme(axis.text.x = element_text(angle = 90, hjust = 1))
# p = p + facet_grid(.~cogdx)
# p

p = ggplot(cellprop1, aes(x = Celltype, y = Fraction, color = cogdx)) + geom_boxplot()
p = p + theme(axis.text.x = element_text(angle = 90, hjust = 1))
p
```
### Synapse store
```{r syn.store, cache=TRUE}
CODE = Folder(name = 'MAYO Cell Type Abundance with Reprocessed RNASeq', parentId = parentId)
CODE = synStore(CODE)

cellprop %>%
  rownameToFirstColumn('SampleID') %>%
  write.table(file = 'MAYO_CellFraction_Reproc.tsv', row.names = F, sep = '\t', quote = F)
obj = File('MAYO_CellFraction_Reproc.tsv', name = 'MAYO Cell Fraction', parentId = CODE$properties$id)
obj = synStore(obj, activityName = activityName, activityDescription = activityDescription,
               used = ALL_USED_IDs, executed = thisFile)
```
